---
title: "Nelder-Mead Algorithm"
author: "Chris Nurrenberg"
date: "September 17, 2018"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Rainfall as a secondary effect of hurricanes is often unpredictable on a yearly basis, considering that hurricane trajectory itself is harder to predict. One particular model used for describing the distribution of the rainfall is the Gamma distribution. The parameters are estimated using the maximum likelihood method over the data points. The formula has no closed-form derivative, so a numberic approach to the solution is required. The objective in this portfolio is to reimplement an existing algorithm (Nelder-Mead) to estimate the parameters for the Gamma distribution.

## Methodology
The function <code>GammaMLE</code> required two inputs: x, the input data; and maxIter, the maximum number of iterations to run in the algorithm.

Before the main algorithm is defined, a helper function for the likelihood calculation is necessary. Using the dgamma function with the log=TRUE option eliminated the need for manually hardcoding the maximum likelihood function. 

```{r, eval=FALSE}
likelihood <- function(x, abVec){
  # x - datapoints to apply the maximum likelihood function to
  # abVec - a vector containing (alpha, beta) for the gamma distribution
  # exponentiate the input for domain restrictions
  abVec <- exp(abVec)
  sum <- sum(dgamma(x, abVec[1], scale=abVec[2], log=TRUE))
  return(-1 * sum)
}

```
With the helper function created, outlining the general algorithm can continue. The first calculation generates appropriate initial estimates of the paramters of the gamma distribution.
```{r, eval = FALSE}
xBar <- mean(x)
beta <- xBar/((sum(x^2)/length(x))-xBar^2)
alpha <- xBar * beta
```
Then, the algorithm commences with three vectors of parameters to optimize. Since we are generating the parameter estimates directly, the logarithm of the values gives the starting parameter to optimize.

The data structure used is a 3D-array to store the vectors at each iteration. The indexing scheme used allows for conveniently storing the results of each iteration as a vector of vectors.
```{r, eval = FALSE}
  tMatrix <- array(rep(0, 2 * 3 * maxIter), c(2,3,maxIter))
# example of storing results. t1, t2, t3 are vectors of the form (alpha, beta)
tMatrix[,,1] <- c(t1, t2, t3)
```
Another matrix was used to store the likelihood results of each iteration. A memory overhead was chosen to increase performance. A similar indexing scheme was used so built-in assignment organized the values correctly.
```{r, eval = FALSE}
# likelihood(theta) matrix
ltMatrix <- array(rep(0, 3 * maxIter), c(3,maxIter))
ltMatrix[,1] <- c(lt1, lt2, lt3)
```

After initial calculations and preparations have been performed, the rest of the stages of the algorithm begin. An iteration count check is performed to prevent infinite iteration, and a tolerance check is performed when insignificant progress is being made. The Ordering Step gives ordered indexing of each paramater vector based on the result from the maximum likelihood function. The indices are stored for later use.

```{r, eval = FALSE}
# Ordering Step
t <- which.max(ltMatrix[,currentIter])
r <- which.min(ltMatrix[,currentIter])
# resultant median is sum of indices minus prescribed indices
s <- 6 - r - t
```

The Centroid step. The center betweent the points with the highest likelihood score is created.

```{r, eval = FALSE}
# Centroid Step, c = (r + s) / 2
cVec <- (rVec + sVec)/2
```

The Reflection Step. The factor of reflection was chosen to be 1 for this implementation.

```{r, eval=FALSE}
# Reflcetion Step
hVec <- cVec + 1 * (cVec - tVec)
# l(r) < l(h) < l(s)
if (ltMatrix[r, currentIter] < likelihood(x, hVec) & likelihood(x, hVec) < ltMatrix[s, currentIter]){
  # c(r, s, h)
  tMatrix[,,currentIter + 1] <- c(rVec, sVec, hVec)
}
```

The Expansion Step. An expansion factor of 2 was chosen for this implementation.

```{r,eval = FALSE}
# Expansion Step
iVec <- cVec + 2 * (hVec - cVec)
if (likelihood(x, iVec)< lh){
  tMatrix[,,currentIter + 1] <- c(rVec, sVec, iVec)
} else {
  tMatrix[,,currentIter + 1] <- c(rVec, sVec, hVec)
}
```

And, finally, the Contraction / Shrinking step. The contraction / shrinking factor was 1/2 for this implementation.

```{r, eval=FALSE}
# Contraction Step
jVec <- cVec + (tVec - cVec)/2
if (likelihood(x, jVec) < ltMatrix[t, currentIter]){
  tMatrix[,,currentIter + 1] <- c(rVec, sVec, jVec)
} else {
  # Shrink Step
  tMatrix[,,currentIter + 1] <- c(rVec + (tMatrix[,1,currentIter] - rVec)/2,
                                rVec + (tMatrix[,2,currentIter] - rVec)/2,
                                rVec + (tMatrix[,3,currentIter] - rVec)/2)

```

After the appropriate steps are applied, the iterative process repeats until it converges, or iterates beyond the supplied limit.

## Results

The algorithm applied to the supplied data resulted in an alpha value of 2.187, and a beta value of 3.331. The expected value of the function outputs .656 inches of rain.

Considering the distribution of the input data, this result does not make sense. Something has failed in the algorithm, and I have not been able to find the error after hours of searching. I had a few errors earlier on in the development process that included using indices over the respective vectors. I've tested my indexing scheme multiple times to make sure nothing is being stored incorrectly. I have not been this genuinely stumped on finding an error in months

##Appendix

```{r}
GammaMLE <- function(x, maxIter=500){
  #grab initial alpha and beta as estimates
  xBar <- mean(x)
  beta <- xBar/((sum(x^2)/length(x))-xBar^2)
  alpha <- xBar* beta
  print(c(alpha, beta))
  #t for theta
  t1 <- log(c(alpha, beta))
  t2 <- t1 + c(t1[1], 0)
  t3 <- t1 + c(0, t1[2])
  # theta[k, i] indexes to 
  #setup theta 3D array, [alpha/beta, i, k]
  tMatrix <- array(rep(0, 2 * 3 * maxIter), c(2,3,maxIter))
  tMatrix[,,1] <- c(t1, t2, t3)
  
  
  lt1 <- likelihood(x, tMatrix[,,1])
  lt2 <- likelihood(x, tMatrix[,,2])
  lt3 <- likelihood(x, tMatrix[,,3])
  
  # likelihood(theta) matrix
  ltMatrix <- array(rep(0, 3 * maxIter), c(3,maxIter))
  
  ltMatrix[,1] <- c(lt1, lt2, lt3)
  currentIter <- 1
  
  while (currentIter < maxIter){
    # Terminating Check
    if (sd(ltMatrix[,currentIter]) < 10^-8){
      break
    }
    # Ordering Step
    t <- which.max(ltMatrix[,currentIter])
    r <- which.min(ltMatrix[,currentIter])
    # resultant median is sum of indices minus prescribed indices
    s <- 6 - r - t
    
    
    #assignment of vectors for later use
    tVec <- tMatrix[,t,currentIter]
    rVec <- tMatrix[,r,currentIter]
    sVec <- tMatrix[,s,currentIter]
    # Centroid Step
    cVec <- (rVec + sVec)/2
    # Reflcetion Step
    hVec <- cVec + (cVec - tVec)
    # Conditional Check
    lh <- likelihood(x, hVec)
    b1 <- ltMatrix[r,currentIter] < lh
    b2 <- lh < ltMatrix[s, currentIter]
    if (b1 & b2){
      # c(r, s, h)
      tMatrix[,,currentIter + 1] <- c(rVec, sVec, hVec)
    } else if (lh < ltMatrix[r,currentIter]) {
      # Expansion Step
      iVec <- cVec + 2 * (hVec - cVec)
      if (likelihood(x, iVec)< lh){
        #c(r,s,i)
        tMatrix[,,currentIter + 1] <- c(rVec, sVec, iVec)
      } else {
        #c(r,s,h)
        tMatrix[,,currentIter + 1] <- c(rVec, sVec, hVec)
      }
      
    } else if (!b2) {
      # Contraction Step
      jVec <- cVec + (tVec - cVec)/2
      
      if (likelihood(x, jVec) < ltMatrix[t, currentIter]){
        tMatrix[,,currentIter + 1] <- c(rVec, sVec, jVec)
      } else {
        # Shrink Step
        tMatrix[,,currentIter + 1] <- c(rVec + (tMatrix[,1,currentIter] - rVec)/2,
                                        rVec + (tMatrix[,2,currentIter] - rVec)/2,
                                        rVec + (tMatrix[,3,currentIter] - rVec)/2)
      }
    }
    
    #
    currentIter <- currentIter + 1
    ltMatrix[,currentIter] <- c(likelihood(x, tMatrix[,1,currentIter]), 
                               likelihood(x, tMatrix[,2,currentIter]),
                               likelihood(x, tMatrix[,3,currentIter]))
  }
  finalIndex <- which.min(ltMatrix[,currentIter])
  print(exp(tMatrix[,finalIndex,currentIter]))
  return(exp(tMatrix[,finalIndex,currentIter]))
}

likelihood <- function(x, abVec){
  # abVec is c(alpha, beta), for conveniently calling from the main algorithm

  abVec <- exp(abVec)
  prod <- sum(dgamma(x, abVec[1], scale=abVec[2], log=TRUE))
  return(-1*prod)
}
y <- c(31.00, 2.82, 3.98, 4.02, 9.50, 4.50,
       11.4, 10.71, 6.31, 4.95, 5.64, 5.51,
       13.4, 9.72, 6.47, 10.16, 4.21, 11.60,
       4.75, 6.85, 6.25, 3.42, 11.80, .8,
       3.69, 3.10, 22.22, 7.43, 5.00, 4.58,
       4.46, 8.00, 3.73, 3.50, 6.20, .67)
answer <- GammaMLE(y)
answer[1]/answer[2]
```